Visual narratives of the Covid-19 pandemic

### 2022-08-31 Review/Response.qmd

---
title: "Reviewer Response"
output: html_document
---

# Main Issues

## Reviewer 1

::: callout
### How was the sample of publications selected? Under what criteria?

As it stands, the selection seems arbitrary and very limited. Why the Financial Times and The New York Times and not others? Why not analyzing visualizations from non-Western countries? If the authors want to talk about a “global narrative”, the selection of graphics should also be global. A comparative cross-cultural study would be pretty productive, as it might reveal similarities and differences of graphics from different regions.
:::

The authors are from the United States and Germany; while we participated in many discussions on global social media platforms such as Twitter, we do not have the global perspective (or even facility with sufficient languages) to undertake a review of a global sample of media relating to COVID.

::: callout
### Principles

The authors say that Edward Tufte’s optimization of data-to-ink ratio is a “general principle of visualization”. That’s quite an overstatement. Tufte’s data-to-ink ratio is more of a very subjective, highly flexible guideline that is extremely dependent on factors such as the type of audience, culture, publication, etc.

To begin with, empirical evidence doesn’t support at all that highly stripped-down, high data-to-ink graphics support understanding in any way. As long as the non-data ink in a visualization doesn’t grossly obscure the representation of the data itself, it’s harmless. Moreover, it can even be beneficial, as it might help attract a reader’s attention to the information —in some studies, readers expressed their dislike for the highly sanitized, modernist/minimalist visual style Tufte is an advocate for.

I’d eliminate the reference to this non-principle and I’d stick to the other principle, which is truly advisable and normative; from the paper itself: “ensuring clear understanding by organising the graphics in such a way that the story of the data is told most effectively.”

This would need to be extended, though, and more clearly explained. Making a visualization understandable doesn’t depend just on organization. It also involves making choices about encodings, types of charts and maps and, more importantly (something that I think this paper overlooks) actually testing that the graphic actually conveys the intended information to the audience it’s supposed to inform. Anyone who works in visualization will tell you that simply applying “good design practices” to a visualization will never guarantee for certain that it is understood.

:::

We have modified the principles section to remove the reference to Tufte and the data/ink ratio. We have slightly reorganized the introduction section to add subsections, and have added an additional paragraph which discusses the series of choices made in the process of designing a data visualization. While the general guidelines for making these choices are somewhat outside the scope of this paper, as this is an article, and not a book, and proper treatment of these decisions in a broader context would require at least a book, we hope that highlighting the more general context surrounding the creation of data visualizations and explicitly defining the scope for this article provides some additional clarity.

::: callout
### Graphics Commentary

I believe that the commentary on the graphics the authors chose would need to be more extensive and detailed. Just to give an example, they compare choropleth maps to proportional symbol maps, but they don’t reflect on the many shortcomings of the latter, such as the usual overlap between the scaled symbols. This might be related to space limits in the journal, but I think that this is a paper that deserves to be at least double its length to be truly comprehensive and detailed enough.

The authors also make too many assumptions about how a graphic will be interpreted. This is a direct quote: “The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data”. Where is the evidence for such an assertion? The fact that the authors —likely professionals who are very familiar with statistics and visualization— find a graphic intuitive doesn’t make that graphic intuitive to other audiences. (The authors themselves actually address this in the following section, about log scales, arguably the best one in the paper.)

:::

While the editor has cautioned us not to double the length of the paper, where possible we have attempted to address some obvious shortcomings in our initial discussion without unduly increasing the paper's length.

## Reviewer 2

::: callout
### Phrasing

This is clearly an article by researchers for whom English is likely a second language. While fluent, the wording is often obtuse and clunky. I have made a number of changes in the text to show where it might be improved.
:::

We appreciate the edits.

:::
### Organization

More importantly, there is a mass of data here and a failure of organization. The whole requires a rewrite that distinguishes with one example for each the use of chloropleth, dot maps, tables, annimations, etc. In the chloropleth section and dot map one can distiguish between raw numbers and those that are, for example, cases/per 1000 persons. The use of sliders and annimation would be a separate section.

The goal is simple. A better abstract would say:  Maps and graphs have been extensively used in presenting the evolving state of the Covid-19 pandemic globally, nationally, and locally. This article seeks to review the types of maps and graphs employed by major news media and to assess the degree to which they served as communicative arguments useful both as a basis of public understanding of the pandemic and, at different stages, the rationale behind policies implimented to reduce its impact.

As is, it reads badly and the organization is such that its points get lost. Keep the illustrations to the points you wish to make. Distinguish betwen map types and their uses. Use A, B-heads as you reorder your argument.

And in the discussion ... what have we learned? Well, there are static and dynamic maps, the former most common in news presentations, the latter in web-based. A trivial but useful point. Sliders are a good example how the latter expands what is in the traditional news map. 
:::

The organization of this paper is certainly not the same as that which is proposed, in part because we have taken a narrative view that explores the different communication narratives of visualizations during the pandemic. Because we focus on the narratives and look at the chart types employed in service of those narratives, organization necessarily deviates from a straightforward analysis of each chart type in sequence, but this is by design.

### 2022-08-31 JDSSV_manuscript_submitted.Rnw
Updates

Over the past two years, several waves of Covid-19 infections with different mutants of the SARS-CoV-2 virus have swept across the globe, claiming many lives, causing numerous health damages and affecting our personal lives in many ways. According to the WHO, over 304 million confirmed cases and over 5.4 million deaths have been reported as of early January 2022 \footnote{\url{https://www.who.int/publications/m/item/weekly-epidemiological-update-on-covid-19---11-january-2022}}. The pandemic has generated enormous interest in epidemiological data, its analysis and visualisation. From the beginning of the pandemic, data on the number of infections and covid-related deaths has been published daily and made available to the public. Media, politicians and individuals use this data to build their narratives about the pandemic, discuss its evolution, justify the measures taken and discuss different prevention strategies against the spread of the virus. So there are several goals to be achieved by visualising Covid-19 data. These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace. But as with any other data visualisation two general principles remain the same: ensuring clear vision by optimising the data-to-ink ratio \citep{Tufte2001} (or signal-to noise ratio) and ensuring clear understanding by organising the graphics in such a way that the story of the data is told most effectively.

In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic as they have been showcasing their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods. Data journalism will certainly be one of the beneficiaries of the Covid-19 pandemic and it has become an innovative part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards.

@@

At the same time, we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org). Numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information have been prepared, see for example \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.

@@

A best practice for effective communication is to present information in an understandable way \citep{gigerenzerHelpingDoctorsPatients2007,gigerenzerWhatAreNatural2011}, for example by saying "one in ten" instead of 10\%. Using absolute rather than relative numbers, presenting information in an appealing graphic form and summarising the most important facts in "fact boxes" are some of the methods that have been developed and advocated to increase transparency in communication between stakeholders such as healthcare providers and patients (see for example: \url{https:// www.hardingcenter.de}). Such "fact boxes" combined with the dynamic layer of the internet allowed for illustrative simulations of the spread of the epidemic, for example as presented already on 14 March 2020 in the Washington Post\footnote{\url{https://www.washingtonpost.com/graphics/2020/world/corona-simulator/}} with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve".  Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge. While human thinking tends to simplify patterns and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate. Thus, in studying COVID-19 and predicting its spread, it is not only important to consider symptomatology, disease incidence and geographic distribution, population behaviour patterns, government policies, and impacts on the economy, schools, people in nursing homes, and society at large, but also to incorporate these into data analyses and communication of results. Associations observed in the data can often be caused by third party variables (confounders). In addition, much of the data comes from observational studies, which usually makes robust causal attribution problematic. However, statisticians who point out these limitations are at risk of having their statements pulled out one-sidedly in a polarised debate \citep{McConway2021}.

@@

On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global distribution of the disease has been in the public eye. The spatial spread of the virus and the resulting cases and deaths are commonly visualized by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022. For the pandemic perspective, a central element of the narrative is the ubiquity of the disease and the accompanying global impact. Choropleth maps based on raw numbers of cases might look convincing and fit to the purpose, but neglect a number of well-know caveats for statistical reporting and visualisation:

\begin{enumerate}
\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers. Especially in the case of similarly coloured areas of the regions, viewers tend to integrate them unconsciously and perceive choropleths as representations of density. They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases. Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997, color2021}. The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.  
\end{enumerate}

@@

Choropleth maps are quite commonly used by media companies and governmental organisations and it is easy to find good and bad examples of their usage. Proportional symbol maps, see Fig.~\ref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region. 
Clearly, the visual impression of any map based illustration depends on the projection chosen for the underlying map as the examples quite often demonstrate a western-centric viewpoint.

@@

The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe. None of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them. The use of maps in this context more often focuses on the comparative aspects: how are we doing as opposed to our neighbors, which strategy to fight Covid-19 is better? We will examine these topics more closely in Section~\ref{sec:rankings}.

@@

\begin{figure}
\includegraphics[width=.99\linewidth]{ft-off-charts}
\caption{Chart showing the increase in cases as a result of the B.1.617.2 variant in regions of the UK. Notably, this chart arranges areas to accommodate certain regions whose cases are ``off the chart" in order to emphasize the exceptional situation. Source: Tweet by John Burn-Murdoch (@jburnmurdoch), designer for the Financial Times. \url{https://twitter.com/jburnmurdoch/status/1397995388267810818}.}
\label{fig:jbm-tweet}
\end{figure}

@@

Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis. Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis. This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.

@@

Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary overview that allowed viewers to make comparisons between regions and points in time without the burden of numerical calculations. Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}. The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is fairly opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate. In addition, by hiding these details from the user, it is very difficult to identify when the representation changes over time, as it did several times during the summer of 2020. The use of a similar color scheme as well as similar representations made it extremely difficult for users to identify that the value represented had changed. Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.

@@

While traditionally, time-series information has been limited to charts with time on the x-axis, there have been many attempts to make use of the capabilities of web graphics and animation in order to convey temporal information in less traditional time-series forms. One controversial chart appeared in the New York Times; the x-axis was wrapped around an Archimedean spiral to provide a sense of periodicity in the year-by-year evolution of case counts, as shown in \Cref{fig:nyt-spiral}. The original form of this chart is prone to the line-width illusion \citep{vanderplasSignsSineIllusion2015}, in addition to the well-known problems with polar charts \citep{hofmannGraphicalTestsPower2012,waldnerComparisonRadialLinear2020}. The re-envisioned version, which aligns counts on the outside of the spiral, and displays them as radial lines, mitigates some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons. In addition, the re-envisioned chart includes a reference line at 100K cases/day, which allows the reader to compare the severity of different waves directly. Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period. In addition, this type of chart devotes less area to previous years of the pandemic, which may decrease the amount of focus given to that data; whether this is an advantage or disadvantage depends on your perspective and goals when using the chart.

@@

Another attempt to creatively show time-series data that has been extremely common online throughout the pandemic is the bar chart race: an animated series of bar charts over time that show how case counts in each country are increasing. A typical example can be found on YouTube at \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}; a screenshot is shown in \Cref{fig:barchart-race}. This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts. By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.

@@

As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time. In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to covid, with occasional short quotes from individuals' obituaries to humanize the names\citep{barryRemembering1000002020}. While this is not a typical time-series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way.

>>>>

Over the past two years, several waves of Covid-19 infections caused by different mutations of the SARS-CoV-2 virus have swept across the globe, claiming lives, stressing healthcare systems, and changing many different facets of our day-to-day experience. According to the WHO, almost 600 million confirmed cases and over 6.45 million deaths have been reported as of late August 2022 \footnote{\url{https://www.who.int/docs/default-source/coronaviruse/who_mou_30_august-2022.pdf}}. The pandemic has generated enormous interest in epidemiological data, its analysis and visualisation. From the beginning of the pandemic, data on the number of infections and covid-related deaths has been published daily and made available to the public. Media, politicians and individuals use this data to build their narratives about the pandemic, discuss its evolution, justify the measures taken and discuss different prevention strategies against the spread of the virus. %So there are several goals to be achieved by visualising Covid-19 data.
These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace, but as with any other data visualisation the general principles remain the same: ensuring clear understanding by organising graphics in such a way that the story of the data is told most effectively.

\subsection{Data Journalism and the Media}

In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods. COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the Covid-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards.

@@

At the same time, we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org). Efforts to improve the communication of quantitative and statistical information date from the beginning mass media and the ability to create charts and tables from data systematically, and while there continue to be efforts made in this area, historical papers outline many of the same issues that we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}. In other words, while our **technology** for recording, visually summarizing, and communicating information distilled from data has vastly improved over the past 100 years, we still face many of the same problems when **designing** charts which communicate data accurately and effectively.

@@

\subsection{Effective Data Communication}
A best practice for effective communication (graphical or otherwise) is to present information in an understandable way \citep{gigerenzerHelpingDoctorsPatients2007,gigerenzerWhatAreNatural2011}, for example by saying "one in ten" instead of 10\%. Using absolute rather than relative numbers. 
Presenting information in an appealing graphic form and summarising the most important facts in "fact boxes" are some of the methods that have been developed and advocated to increase transparency in communication between stakeholders such as healthcare providers and patients (see for example: \url{https:// www.hardingcenter.de}). 
Such "fact boxes" combined with the dynamic layer of the internet allowed for illustrative simulations of the spread of the epidemic, for example as presented already on 14 March 2020 in the Washington Post\footnote{\url{https://www.washingtonpost.com/graphics/2020/world/corona-simulator/}} with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve".  Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge. 
While human thinking tends to simplify patterns and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate. 
Thus, in studying COVID-19 and predicting its spread, it is not only important to consider symptomatology, disease incidence and geographic distribution, population behaviour patterns, government policies, and impacts on the economy, schools, people in nursing homes, and society at large, but also to incorporate these into data analyses and communication of results. 
Associations observed in the data can often be caused by third party variables (confounders). 
In addition, much of the data comes from observational studies, which usually makes robust causal attribution problematic. 
However, statisticians who point out these limitations are at risk of having their statements pulled out one-sidedly in a polarised debate \citep{McConway2021}.

\subsection{Data Visualization Design and Testing}
Having briefly discussed the media landscape surrounding the COVID pandemic and goals for communicating about data effectively, we should also briefly discuss the choices that go into creating visual representations of data intended to convey quantitative information.
While the overall goal of a data visualization is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what forms are used to represent the data? 
How will the chart (and any surrounding story, in journalism) be presented to the viewer? 
As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualization (computer, physical newspaper, tablet, cell phone) - all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.

Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organizations, but may also be deliberately chosen for visual effect: color palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyze in this paper did not simply pop into being - they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
We will highlight particular choices which were effective (or not) throughout this paper, but we will also leave some undiscussed for the sake of brevity.

Of course, the creative process does not only involve the creator. Most graphics that are created for a purpose are tested, either formally or informally, before they are published, even if the test results are never themselves published. Because so many of the charts we highlight in this paper were discussed and disseminated via social media, in some cases we can provide contemporary discussions of the design process and reactions to the charts we highlight in this paper. Where such examples are not provided, that should in no way be taken as evidence that the design and testing process was less formal; it may only have been less public.

\subsection{Scope}
Our focus in this paper is to highlight, analyze, and discuss a selection of charts and graphics which caught our attention during the initial, real-time experience of the pandemic. As we primarily consume media from western Europe and the United States, our assessment is primarily limited to news outlets catering to these audiences; a retrospective global analysis of COVID graphics would certainly be interesting, but is beyond the scope of this work. While the charts we have selected were of personal interest during the fear and uncertainty of the initial stages of the COVID pandemic, in this paper we assess these same charts retrospectively, from a world where COVID is an acknowledged fact of life. While it still has significant impacts on our day-to-day lives, the existence of treatments, vaccines, and other preventative measures, as well as the human tendency to adapt to changing circumstances mean that the novel coronavirus which emerged in 2019 no longer dominates our consciousness in the same way. Because of this perspective difference, we have enough distance from the initial cataclysm to be able to assess the emotional impact of the graphics at the time and to assess their utility retrospectively.

@@

On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global distribution of the disease has been in the public eye. The spatial spread of the virus and the resulting cases and deaths during this initial period were commonly visualized by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022. A central element of the narrative during March 2020 was the ubiquity of the disease and the accompanying global impact. While choropleth maps based on raw numbers of cases might look convincing and are an obvious choice under the circumstances, the use of these charts ignores a number of well-know caveats for statistical reporting and visualisation:

\begin{enumerate}
\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers. Viewers tend to integrate similarly coloured areas of the map unconsciously, and perceive choropleths as representations of density. They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases. Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
\item Colour-scheme selection: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997, color2021}. The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.
\end{enumerate}

@@

The issues with choropleth maps were highlighted in trade publications during the initial stages of the pandemic \citep{fieldMappingCoronavirusResponsibly}, but these publications did not penetrate the bureaucracy of public health officials suddenly tasked with data visualisation in realtime at scale. 
Choropleth maps are quite commonly used by media companies and governmental organisations and it is easy to find good and bad examples of their usage.
Proportional symbol maps, see Fig.~\ref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. 
The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region. 
Proportional symbol maps sacrifice familiar shapes and topological connectedness for equal visual weight\citep{gaoUsabilityValuebyalphaMaps2019}; while this representation avoids the area bias, proportional symbol maps are subject to under-estimation biases \citep{shimAnalysisMiddleSchool2008} that may be of particular concern when visualizing epidemic disease spread.
In addition, the visual impression of any map based illustration depends on the projection chosen for the underlying map; the examples in \Cref{fig:choro1} and \Cref{fig:propsymb} have a western-centric bias.

@@

Another modification of the standard choropleth to address the area-bias problem is to relax the spatial borders of a region in order to show magnitude using area (as well as color). Various algorithms can be used to modify geographic boundaries to maintain spatial continuity, or shapes such as hexagons or squares can be used to represent geographic entities (as in \Cref{fig:hex-cartogram}); with either choice, geometry, topology, or both must be sacrificed.  While these methods are more popular in the academic visualisation literature, including analysis of the spread of COVID-19\citep{yalcinMappingGlobalSpatiotemporal2022}, they were not overly common in the media for global representations of COVID information. This may be because cartograms require more visual effort on the part of the viewer; in addition, news organizations may be reluctant to assume their readers have the geographic sophistication necessary to translate between the distorted map of the globe as drawn on the cartogram and the undistorted mental map of the globe used for mental comparison purposes. 
A spatial dot density map might have allowed for analysis of data across geographic boundaries, but this approach was not commonly used to show global case counts, though it was highlighted as a useful option at the national level \citep{fieldMappingCoronavirusResponsibly}, as shown in \Cref{fig:china-dot-map}.

The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe: none of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them. 
The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight Covid-19 is better? 
We will examine these topics more closely in Section~\ref{sec:rankings}.

\begin{figure}
\centering
\begin{subfigure}[c]{.45\textwidth}
\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state. The COVID Tracking Project \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
\end{subfigure}\hfill
\begin{subfigure}[c]{.45\textwidth}
\includegraphics[width=\textwidth]{china-dotmap}
\caption{A dot-density map of COVID cases in China \citep{fieldMappingCoronavirusResponsibly}.}\label{fig:china-dot-map}
\end{subfigure}
% \begin{subfigure}[t]{.45\textwidth}
% \includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
% \caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
% \end{subfigure}
\caption{Modifications of choropleth maps which attempt to address the area bias of choropleths in different ways. Both methods are applied at the national scale rather than at global scale because of limitations with the visual form or data quality.}
\end{figure}

@@

\begin{figure}
\includegraphics[width=.99\linewidth]{ft-off-charts}
\caption{Chart showing the increase in cases as a result of the B.1.617.2 variant in regions of the UK, May 27, 2021. Notably, this chart arranges areas to accommodate certain regions whose cases are ``off the chart" in order to emphasize the exceptional situation. \citep{burn-murdochNEW1617Fuelling2021}}
\label{fig:jbm-tweet}
\end{figure}

Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex: over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}. Interestingly, media outlets and municipal dashboards were subject to criticism when the color scales changed in response to increasing case counts\citep{boeckDataVizDecisionmakingAction2020} as well as when the color scale did not change\citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualizations amid competing constraints and a constantly evolving pandemic.

@@

Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis. 
Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig-jbm-tweet} and \Cref{fig:wapo-covid-time-series}. 
This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.

@@

Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary overview that allowed viewers to make comparisons between regions and points in time without the burden of numerical calculations. Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}. The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is fairly opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate. In addition, by hiding these details from the user, it is very difficult to identify when the representation changes over time, as it did several times during the summer of 2020. The use of a similar color scheme as well as similar geometric representations made it extremely difficult for users to identify that the value represented had changed. Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.

@@

While traditionally, time-series information has been limited to charts with time on a linear x-axis, many charts manipulated time to highlight different aspects of the pandemic. In some cases, these modifications provided large amounts of utility for a relatively small increase in complexity: for example, John Burn-Murdoch's decision to index Financial Times covid graphs to the days since the $N$th case or death. This decision is a small modification of the basic time-series plot, but has important consequences in the chart's readability as well as the visual emphasis for users: the graphs in the Financial Times were designed to emphasize "... that there's an inevitability about how coronavirus spreads... even if there are only a few cases in your country today, based on all the data we have, you will end up going along that path, the same path that the likes of Italy and Spain have been on so tragically."\citep{hannenCoronavirusTrajectoryTracker2020a} By placing all countries on an even footing divorced from the temporal randomness of when the first case appeared, the viewer can effectively compare the effect of government policies and other interventions on the trajectory over time\citep{burn-murdochAllCountriesDeath2020,burn-murdochLatestCaseTrajectories2020}.

The New York Times intentionally violated typical design guidelines in a controversial chart; the x-axis was wrapped around an Archimedean spiral to provide a sense of periodicity in the year-by-year evolution of case counts, as shown in \Cref{fig:nyt-spiral}. The original form of this chart is prone to the line-width illusion \citep{vanderplasSignsSineIllusion2015}, in addition to the well-known problems with polar charts \citep{hofmannGraphicalTestsPower2012,waldnerComparisonRadialLinear2020}. A re-envisioned version created by Sourya Shrestha, aligns counts on the outside of the spiral, and displays them as radial lines, mitigating some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons. In addition, the re-envisioned chart includes a reference line at 100K cases/day, which allows the reader to compare the severity of different waves directly. Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period. By design, this type of chart devotes less area to previous years of the pandemic, which may decrease the amount of focus given to that data; whether this is an advantage or disadvantage depends on your perspective and goals when using the chart.

@@

Other designers used the capabilities of web graphics to show the time component in the data through animation. In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race": an animated series of bar charts over time that show how case counts in each country are increasing. A typical example can be found on YouTube at \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}; a screenshot is shown in \Cref{fig:barchart-race}. This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts. By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.

@@

As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time. In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanize the names\citep{barryRemembering1000002020}. While this is not a typical time-series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way. A less sentimental dot-density timeline by the New York Times about 9 months later chronicled the nation reaching 450,000 deaths; the difference between the two visualizations brings to mind the quote oft attributed to Stalin: "One death is a tragedy; a million deaths is a statistic" \citep{investigatorSingleDeathTragedy2016}.
